Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0

### 2019-03-02 manuscript.Rmd
updates

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across 44 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye tissue RNA-seq samples. The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote. The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing quick visualizations by database version, gene/transcript, and 60 tissues. As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but contain several gene set differences that suggest organoid process improvements. CONCLUSION: The eyeIntegration v1.0 web app serves the pan-eye and body transcriptome dataset, EiaD. This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
---

@@

## The GTEx gene expression resource lacks eye-specific tissues
The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017]. However, no eye tissues are included in their resource. The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018]. To date, no curation of the sample level metadata has been done, so it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed. More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016]. iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only.

**ADD MENTION OF NEW SWAROOP DATA "GREx"**

## eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018]. This interactive web app allows for quick transcript and gene comparisons across many eye tissues and dozens of other body tissues. The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to hand-assess quality control for the samples. The interactive nature of some of the steps precludes efficient and regular data updates for the data.

To better meet the needs of the eye research community we have re-written the bioinformatic pipeline that creates the eye and body RNA-seq dataset to allow for regular, versioned updates for eyeIntegration. We call this reproducible and versioned transcriptome dataset "Eye in a Disk" (EiaD). The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provide consistent sample quality control, and improves efficiency for future sample updates. The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples and improved sample labeling. The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data tables searching, easy download of core datasets, and local install of the entire app with three commands. Additionally, we are prototyping new tools to display single cell RNA-seq (scRNA-seq) data to provide researchers access to cell type specific information about gene expression across murine retinal development.

## EiaD dataset and eyeIntegration used to identify potential avenues to improve retina organoid maturation

Retina organoids

@@

Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue. Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC). Another substantial addition to the 2019 EiaD are non-protein coding genes. We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).

@@

## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups
We performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing quick identification of eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons. We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues. The user can view the results selecting ‘Differential’ under the ‘Expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).  As with 'Expression', the user can select which version of the web app to draw data from as well as select for gene- or transcript-level comparisons. The user additionally has the option to select different gene classes to examine, e.g. protein coding, lincRNA.

@@

We use our large dataset to narrow in on some processes which to identify three core processes which differ. First we show that the homebox (HOXB) family is overexpressed in the organoids. The homeobox family is well known to initiate polarity of the embryo during early development. As retinoic acid, which interacts directly to control expression of HOX genes, is used in the differentiation process from stem cell to organoids, it is likely that the HOXB expression is a by-product of the application of retinoic acid. While higher HOXB expression may be important to spur differentiation of the multiple cell types of the retina, the continued HOXB expression in the older organoids may inhibit the maturation of the organoid.

Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids. Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival. We would predict that decreased protocadherin expression would reduce RGC activity, which is exactly the result. Many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues. This result suggests that modifying stem cells to increase protocadherin expression may result in higher RGC activity.

https://www.ncbi.nlm.nih.gov/pubmed/30266927

http://dev.biologists.org/content/135/24/4141?ijkey=2ce65c0994c48221620a90803e1861685b60935c&keytype2=tf_ipsecsha

http://jcs.biologists.org/content/128/8/1455

>>>>

---
title: 'Eye in a Disk: eyeIntegration human pan-eye and body transcriptome database version 1.0'
author:
  - Vinay Swamy:
      institute: ogvfb
  - David McGaughey:
      institute:
        - ogvfb
      correspondence: "yes"
      email: mcgaugheyd@mail.nih.gov
institute:
  - ogvfb: Ophthalmic Genetics and Visual Function Branch, National Eye Institute, National Institutes of Health
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  word_document:
    reference_docx: word-styles-reference-01.docx
    fig_caption: yes
    pandoc_args:
      - '--lua-filter=scholarly-metadata.lua'
      - '--lua-filter=author-info-blocks.lua'
bibliography: eyeIntegration_v1_app_manuscript.bib
csl: investigative-ophthalmology-and-visual-science.csl
abstract: "PURPOSE: To provide the eye community with an easily accessible and reliable RNA-seq based transcriptome expression database of healthy human eye tissues and a matching reactive web application to query gene expression in eye and body tissues. METHODS: We downloaded the raw sequnce data for 892 RNA-seq based samples across 44 tissues in the GTEx project as a non-eye reference set. We then queried several public repositories to find all healthy, non-perturbed, human eye tissue RNA-seq samples. The 1311 total samples we found are sent into a Snakemake-based reproducible pipeline we wrote. The pipeline automatically quantifies all known transcripts and genes, removes samples with poor sequence quality and mislabels, normalizes expression values across each tissue, performs 105 differential expression tests, calculates GO term enrichment, and outputs all as a single SQLite database file: the Eye in a Disk (EiaD) dataset. RESULTS: As 40+ GB database is not a useable format for most, we rewrote the web application  eyeIntegration (https://eyeIntegration.nei.nih.gov) to display these data, focusing on providing quick visualizations by database version, gene/transcript, and 60 tissues. As a demonstration of the value of this unified pan-eye dataset, we show that fetal and organoid retina tissues are highly similar at a pan-transcriptome level but display distinct differences in protocadherin and HOXB family members. CONCLUSION: The eyeIntegration v1.0 web app serves the pan-eye and body transcriptome dataset, EiaD. This provides the eye community a powerful and quick mean to testing hypothesis relating to human gene and transcript expression across 44 body and 16 eye tissues."
keywords: "RNA-seq, lens, cornea, retina, RPE, Snakemake, web, app <- Not used in Word as far as I can tell"
---

@@

## The GTEx gene expression resource lacks eye-specific tissues
The Genotype-Tissue Expression (GTEx) Project has generated RNA-seq data across dozens of post-mortem human tissues from hundreds of unique donors and presents the gene and transcript level data in a efficient web app (https://gtexportal.org/) [@carithersNovelApproachHighQuality2015; @gtexconsortiumGeneticEffectsGene2017]. Recently Ratnapriya et al. have published a huge set of post-mortem retina, both normal and with varying degrees of age-related macular degeneration and the GTEx project is providing the data as a download link. This data, as of March 2019, is not available in the interactive GTEx visualizations [@ratnapriyaRetinalTranscriptomeEQTL2019]. The Sequence Read Archive (SRA) and European Nucleotide Archive (ENA) are the primary repositories for all raw sequence data and two groups have quantified large portions of the RNA-seq data, including some human eye tissues, from the SRA: recount2 and ARCHS4 [@collado-torresReproducibleRNAseqAnalysis2017; @lachmannMassiveMiningPublicly2018]. To date, no curation of the sample level metadata has been done, so it is challenging to parse out which eye tissues are present and even more difficult to determine whether any samples were chemically or genetically perturbed. More targeted web resources that allow researchers to quickly assess gene expression in eye tissues include iSYTE, EXPRESS, and retina.Tigem.it [@budakExpressDatabaseTranscriptome2018a; @kakranaISyTEDatabaseExpressionbased2018; @pinelliAtlasGeneExpression2016]. iSYTE only includes lens samples, EXPRESS is limited to a subset of mouse lens and retina samples, and retina.Tigem.it is retina only.

## eyeIntegration app interactivey serves huge GTEx and human eye tissue datasets (EiaD)
The eyeIntegration web resource (https://eyeIntegration.nei.nih), originally released in 2017, provides the largest set of transcriptomes from hand-curated human eye tissues along with hundreds of GTEx tissue samples [@Bryan_2018]. This interactive web app allows for quick transcript and gene comparisons across many eye tissues and dozens of other body tissues. The dataset that the original eyeIntegration web app served was created with a series of scripts, several of which were run interactively to hand-assess quality control for the samples. The interactive nature of some of the steps precluded efficient and regular data updates for the data.

To better meet the needs of the eye research community we have re-written the bioinformatic pipeline that creates the eye and body RNA-seq dataset to allow for regular, versioned updates for eyeIntegration. We call this reproducible and versioned transcriptome dataset "Eye in a Disk" (EiaD). The pipeline automates the EiaD creation, ensures full reproducibility of the results, allow for external data comparison, provides consistent sample quality control, and improves efficiency for future sample updates. The 2019 EiaD dataset contains several new tissue types, full gene product quantification, along with over 100 new samples and improved sample labeling. The eyeIntegration web app has also been re-written to provide many new features, including versioned EiaD datasets, custom URL shortcut creation, new visualizations, improved data tables searching, easy download of core datasets, and local install of the entire app with three commands. Additionally, we are prototyping new tools to display single cell RNA-seq (scRNA-seq) data to provide researchers access to cell type specific information about gene expression across murine retinal development.

## EiaD dataset used to identify potential avenues to improve retina organoid maturation

Retina organoids are an increasingly popular means to model human retina development. We use our pan-study EiaD dataset to demonstrate that, at a pan-transcriptome level, organoids are highly similar to fetal retina tissue. We also demonstrate that important temporal gene expression patterns in the fetal retina tissue is recapitulated in the organoids. As the organoid differentiation methods do not yet produce fully mature retina, we focused on identifying differentially expressed processes, identifying protocadherin and HOXB family gene expression differences which suggest means to improve and benchmark organoid differentiation methods.

@@

Stem cell-derived cornea, stem cell-derived lens, and fetal retina are three new types of sub-tissues that are now available in EiaD. We have also substantially improved the granularity of the cornea tissue metadata, now delineating whether the tissue is from the endothelium or epithelium (`r fig_cap(name = 'tissueCountsBarPlot', display = 'cite')`); previously these had been grouped together as adult tissue. Similarly we more precisely delineate our retina tissues; we drop the "stem cell" sub-tissue classification for organoid and retinal ganglion cells (RGC). Another substantial addition to the 2019 EiaD are non-protein coding genes. We now quantify expression across `r gene_anno %>% count(gene_type) %>% as_tibble() %>% nrow()` gene and  `r tx_anno%>% count(transcript_type) %>% as_tibble() %>% nrow()` transcript types, including protein coding, retained intron, lincRNA, antisense, and pseuodgenes (`r supTab_cap(name = 'tx_and_gene_types', display = 'cite')`).

@@

## Differential expression and gene ontology enrichment tests allow quick comparison of gene differences between groups

We performed multiple differential comparisons at the sub-tissue level within all eye tissues and against a pan-body synthetic set comprised of a stratified sample of all tissues present in our subset of the GTEx dataset, allowing quick identification of eye specific genes across `r limma_DE_tests %>% nrow()` different comparisons. We have expanded the differential tests in the 2019 EiaD by adding the GTEx tissues as direct comparisons to our eye sub-tissues. The user can view the results selecting ‘Differential’ under the ‘Expression’ tab (`r paste0(fig_cap(name = 'appScreenshots', display = 'cite'),'F')`).  As with 'Expression', the user can select which version of the web app to draw data from as well as select for gene- or transcript-level comparisons. The user additionally has the option to select different gene classes to examine, e.g. protein coding, lincRNA.

@@

We use our large dataset to narrow in on some processes which to identify three core processes which differ. First we show that the homebox (HOXB) family is overexpressed in the organoids. The homeobox family is well known to initiate polarity of the embryo during early development [@philippidouHoxGenesChoreographers2013]. Retinoic acid is applied at about day 20 to help differentiate stem cell to organoids and is also known to activate the HOXB family. The lack of HOXB expression at any age in our fetal tissue retina dataset and the broad chromatin and gene expression changes HOXB family members can mediate suggests that HOXB activity may be unwanted for organoid maturation.

Next we demonstrate that many protocadherins are more highly expressed in the fetal tissue, relative to the organoids. Protocadherins mediate cell to cell connections and, in the developing mouse, are shown to be important for spinal internneurons and RGC survival [@hayashiEmergingRolesProtocadherins2015; @lefebvreGProtocadherinsRegulateNeuronal2008]. We would predict that decreased protocadherin expression would reduce RGC activity. Indeed we see that many canonical RGC markers, while present in detectable levels in the organoids, are signficantly underexpressed in the organoids relative to the tissues. This result suggests that modifying stem cells to increase protocadherin expression may result in higher RGC activity.
